---
title: "iBAQ study"
author: "Benjamin Sanchez"
output:
  html_document:
    toc: true
    toc_depth: 4
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Loading and pre-processing data

#### Load packages

```{r warning = FALSE, message = FALSE, results = FALSE}
library(knitr)
```

```{r echo=FALSE}
read_chunk('loadData.R')
read_chunk('combineData.R')
read_chunk('plotData.R')
```

#### Load UPS2

The commercial kit brings 10.6 ug of protein, however only 2.2 ug were used:

```{r loadUPS2}
```

#### Load iBAQ data

In the iBAQ data (IS+ES) there are 6 samples: 3 technical replicates (`batch1`, `batch2` & `batch3`) and each of them processed with 2 different MS methods `top5` & `top10`. We will have then 6 different ES curves:

* top5_batch1
* top5_batch2
* top5_batch3
* top10_batch1
* top10_batch2
* top10_batch3

```{r loadIBAQdata}
```

#### Get ES data

In each sample of the iBAQ data there are:

* 12 ug of IS: yeast samples, all marked; i.e. will appear in the heavy fraction (H)
* 2.2 ug of ES: universal protein standard (UPS2) unmarked; i.e. will appear in the light fraction (L)

```{r getESdata}
```

#### Load SILAC data

There are 18 different samples:

* 3 biological replicates (`R1`, `R2` & `R3`)
* each with 3 technical replicates (`batch1`, `batch2` & `batch3`)
* each estimated with a different MS method (`top5` & `top10`)

Each sample consists of:

* 15 ug of IS (detected in the H fraction)
* 15 ug of actual sample (detected in the L fraction)

```{r loadSILACdata}
```

## 2. Computing absolute abundance of samples

#### Getting IS abundances from ES

Abundance of the IS can be computed by building a log-log standard curve for the ES and applying it to the H peaks in the iBAQ sample. Note that said ES curve MUST have slope = 1, as the underlying assumption is that iBAQ peaks are linearly proportional to the mass of the corresponding proteins, therefore:

`mass = m*iBAQ`

`log10(mass) = log10(m*iBAQ)`

`log10(mass) = log10(iBAQ) + log10(m)`

So if the fit is done in the log space, we fix the ES curve to always have slope = 1, and find the corresponding fit for the intercept:

```{r getISabundance}
```

Let's take a look at the total detected protein amount in each of the 6 samples:

```{r plotTotalProt}
```

```{r fig.align="center"}
mean_totProt <- plotTotalProt(iBAQdata,'Abundance.')
```

<center>**Figure 1: Total detected protein amounts in all 6 IS**</center>
<br>

#### Getting Sample abundances from IS

We now use the absolute abundances from the IS (pg/sample) to infer absolute abundances in each sample of the SILAC data (pg/sample), by doing:

`abundance(sample) = (L/H)ratio * abundance(IS)`

Note that the total mass of IS in the iBAQ sample was of 12 ug, whereas the total mass of IS in the SILAC sample was of 15 ug, therefore we need to rescale the abundances.

```{r getSamplesAbundance}
```

```{r}
SILACdata <- getSampleAbundance(SILACdata,iBAQdata,'Abundance.')
```

Let's take a look at the total detected protein amount of each of the 18 samples, colored by the original ES curve used for the calibration:

```{r fig.align="center"}
mean_sample <- plotTotalProt(SILACdata,'Abundance.R..1_')
```

<center>**Figure 2: Total detected protein amounts in all 18 samples**</center>
<br>

The total detected protein seems to vary quite a bit among samples depending on the ES, so let's take a look at the ES data + standard curves. For that we need:

* A function for plotting a linear fit:

```{r plotLM}
```

* A function for plotting the external standard:

```{r plotES}
```

* A function for plotting all 6 external standards (together or separate):

```{r plotAllES}
```

Now we can take a look at all 6 ES curves:

```{r fig.align="center"}
plotAllES(ESdata,0,FALSE)
```

<center>**Figure 3: ES curves for all 6 measurements of UPS2**</center>
<br>

Those fits give us the transformation from light (L) iBAQ intensity of the UPS2 proteins to abundance (pg/sample). But they don't look the best (consider they are in log), many other curves could almost equally well fit that data. What if instead we assume that all proteins in the IS should add up to the average of the totals shown in figure 1? How would the new ES curves look like?

## 3. Rescaling abundances to the IS in iBAQ

The alternative: All H peaks added up together should result in the average total detection from figure 1:

```{r rescaleIS}
```

```{r}
iBAQdata <- rescaleIS(iBAQdata,mean_totProt)
```


Let's check the total detected protein amount to see if we scaled correctly:

```{r fig.align="center"}
mean_rescaled <- plotTotalProt(iBAQdata,'AbundanceRescaled.')
```

<center>**Figure 4: Total detected protein amounts in all 6 IS, rescaled**</center>
<br>

Now we can look at the new fits of the UPS2 data, as we have esentially created a linear model:

`abundance = m*Hpeak`, where `m = (average total abundance)/(sum of all Hpeaks)`

`log(abundance) = log(Hpeak) + log(m)` -> linear model with a = 1 and b = log(m)

```{r fig.align="center"}
Hpeaks <- grep('iBAQ.H.T4h',names(iBAQdata))   #All 6 absolute iBAQ peaks from the IS (H fraction)
plotAllES(ESdata,mean_totProt/colSums(iBAQdata[,Hpeaks], na.rm = TRUE),FALSE)
```

<center>**Figure 5: ES curves for all 6 measurements of UPS2, original (red) Vs rescaled (green)**</center>
<br>

The fits are almost the same, validating our approach. Let's see everything in the same plot:

```{r fig.align="center"}
plotAllES(ESdata,mean_totProt/colSums(iBAQdata[,Hpeaks], na.rm = TRUE),TRUE)
```

<br>
<center>**Figure 6: ES curves for all 6 measurements of UPS2, original (red) Vs rescaled (green)**</center>
<br>

Note that the MS precision is quite poor: the same protein detection can vary almost a full order of magnitude. This supports the idea that an alternative linear fit could replace the original ES curve, as both would be an appoximation anyway.

Looking now at the total detected protein amounts in the samples:

```{r}
SILACdata <- getSampleAbundance(SILACdata,iBAQdata,'AbundanceRescaled.')
```

```{r fig.align="center"}
mean_sampleRescaled <- plotTotalProt(SILACdata,'AbundanceRescaled.R..1_')
```

<center>**Figure 7: Total detected protein amounts in all 18 samples, rescaled**</center>
<br>

We get much more similar total detected protein amounts, showing that the choice of ES curve will have a major effect in the final abundances. Therefore, one should hopefully have more than just one ES curve, as results will be greatly biased by it.

#### Comparing abundances with/without rescaling

We now display the variability of the final abundance data between biological and technical replicates, and with Vs without rescaling. First we define a scatter plot helping function:

```{r plotScatter}
```

We now define a function that plots all possible combinations between replicates. For instance, for biological replicates, the data in the variable's name regarding biological replicate (`.R1.1`, `.R2.1` and `.R3.1`) will be first removed, and then 2 columns will be plotted against each other if the rest of the name matches (meaning it's the same technical replicate but 2 different biological replicates).

```{r plotVariability}
```

Finally, we define a PCA function to see how the samples cluster:

```{r plotPCA}
```

Let's now see the results of those plots:

```{r fig.align="center"}
par(mfrow = c(2,3), mar = c(0, 0, 1, 0) + 0.5, cex = 1)
# Top row: No rescaling
abundance <- SILACdata[,grep('Abundance.R',names(SILACdata))]
plotVariability(abundance,c('.R1.1','.R2.1','.R3.1'),'Biological Variability')
plotVariability(abundance,c('_Batch1','_Batch2','_Batch3'),'Technical Variability')
plotPCA(abundance,'PCA')
# Bottom row: With rescaling
abundance <- SILACdata[,grep('AbundanceRescaled.R',names(SILACdata))]
plotVariability(abundance,c('.R1.1','.R2.1','.R3.1'),'Biological Variability')
plotVariability(abundance,c('_Batch1','_Batch2','_Batch3'),'Technical Variability')
plotPCA(abundance,'PCA')
```

<center>**Figure 8: Comparison of data variability with (bottom row) and without (top row) rescaling. In the variability plots (left and middle) green is a FC < 2, yellow between 2 and 10, and red > 10. In the PCA plots (right), colors refer to technical replicates and shapes to biological replicates.**</center>
<br>

Even though the variability looks similar in both cases, we see a lower median fold change in the technical replicates + a slight separation of the "technical clusters" in the PCA when we use the rescaling. This means that by using rescaled data we achieve lower variability between technical replicates.

In conclusion, by assuming that the iBAQ data should always adds up to a given protein amount, we can reproduce almost exactly the ES curves, and achieve much more consistent results accross samples. The ES curve achieved with UPS2 therefore is only useful to get an estimate of the percentage of detection in the IS data.

## 4. Rescaling abundances to the IS in SILAC



## 5. Questions

1. Are the differences in figure 1 due to pipetting errors and/or MS detection issues? or something else?
2. Is rescaling to the average total amount the best strategy? It requires then for any proteomic study to run several ES curves (in order to get an idea of the variability of the MS machine). Maybe the detection percentage could be estimated in another way? This is the "purest" way of estimating it I could think of, as it relies only on the MS data and not in any external database (PaxDB, uniprot, etc)
3. Are ES curves always constructed with a slope = 1? Not sure if this was a given in the field or people actually fit both a and b in y=ax+b
4. How relevant are these findings? Old news or actually something interesting? Most times I've only seen a single ES curve...
